Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2020-12-31 sources/bmc_template_submission/bmc_article.tex
example section

(also see \cite{koon,xjon,marg,schn,koha,issnic}).

\section*{Results}
Text for this section\ldots

@@

\begin{figure}[h!]
  \caption{Sample figure title}
\end{figure}

>>>>

\section*{Results}
\subsection*{Case Study: The hollies}

A user is interested in phylogenetic relationships within the genus *Ilex*. Commonly
known as "hollies", the genus encompasses between 400-700 living species, and is
the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering
plants.

An online literature review in June 2020 (Google scholar search for "ilex phylogeny")
reveals that there are several published phylogenies showing relationships within
the hollies \cite{cuenoud2000molecular, manen2010history, setoguchi2000intersectional,
selbach2009new}, but only two have data publicly available \cite{gottlieb2005molecular, yao2020phylogeny}.
\cite{gottlieb2005molecular} made original tree and alignment available in
\href{https://treebase.org/treebase-web/search/study/summary.html?id=1091}{TreeBASE study 1091}.
The tree sampling 41 species was also added to
OpenTree's Phylesystem \href{https://tree.opentreeoflife.org/curator/study/edit/pg_2827/?tab=home}{study pg\_2827}
and it has been integrated into OpenTree's synthetic tree
\href{https://devtree.opentreeoflife.org/opentree/opentree12.3@mrcaott68451ott89474/Ilex-theizans--Ilex-dumosa}{here}.

The most recent *Ilex* tree \cite{yao2020phylogeny} is available in
OpenTree's Phylesystem \href{https://tree.opentreeoflife.org/curator/study/view/ot_1984}{study ot\_1984}
and in the DRYAD repository \cite{yao2020dryad}.
With 175 tips, the \cite{yao2020phylogeny} tree is the best sampled phylogeny yet available for
the hollies.

We ran Physcraper on a laptop Linux computer to update an internal transcribed
spacer DNA region (ITS) alignment from \cite{gottlieb2005molecular}, using a local
GenBank database.  BLAST and RAxML analyses ran for 19hrs 45min, with bootstrap
analyses taking an additional 13hrs.
<!-- **MTH: you probably need some details about the hardware, given the fact
hat you are discussing running times** -->
The updated \cite{gottlieb2005molecular} tree (Fig. \@ref(fig:results)) displays
all 41 distinct
taxa from the original study plus 231 new tips, contributing phylogenetic data to
84 additional *Ilex* taxa. The best RaxML tree is 99% resolved, with 25% of nodes
with bootstrap support < 0.1 and 48% nodes with bootstrap support > 0.75.
A large portion of internal branches are negligibly small, with 30 branches
< 0.00001 substitution rate units, from which only 9 have a bootstrap support > 0.75
(Figure 2).
For comparison, \cite{yao2020phylogeny} also contains all 41 distinct taxa from
the original
\cite{gottlieb2005molecular} study,
and contributes phylogenetic data to 134 additional *Ilex* taxa, from which
`r 135-68` are also in updated \cite{gottlieb2005molecular}. While \cite{yao2020phylogeny} also used
ITS as a marker, their GenBank data is not released yet, so Physcraper was unable
to incorporate 68 additional taxa into the analysis. However, Physcraper was able
to incorporate 18 taxa that were not in \cite{yao2020phylogeny}.

@@

\begin{figure}[h!]
  \caption{A) Phylogeny updated with Physcraper from original
  \cite{gottlieb2005molecular} tree in B.
  Tips in original alignment and new tips added with Physcraper are depicted in
  black and red, respectively. Physcraper obtained sequences from the GenBank
  database via local BLAST of all sequences in the original alignment that generated
  tree in B), filtered them following criteria from section "DNA sequence search and
  filtering", aligned them to original alignment using MUSCLE and performed a
  phylogenetic reconstruction using RAxML with 100 bootstraps. B-D conflict analyses
  performed with OpenTree tools.}
\end{figure}
